---
title: 'R Notebook'
output: html_notebook
---

The researchers induced kidney injury in mice using bilateral ischemia-reperfusion injury (Bi-IRI), which essentially cuts off blood flow to both kidneys using clamps, then restores it. This mimics the kind of kidney injury that can occur in surgical or clinical settings.

Tissue and blood were collected at multiple points: a sham control, then 4 hours, 12 hours, 2 days, and 6 weeks after injury. This allowed the team to track the full arc of injury and repair.

In a sham procedure, the mice go through the entire surgical process, they're anesthetized, cut open, and the kidneys are exposed, but the renal clamps are never actually applied. So blood flow to the kidneys is never interrupted, meaning no actual injury occurs.

We are comparing between sham and the 12 hours after surgery. At the 12 hour time point where they measured creatinine and GFR specifically, because that's when kidney function loss is most clearly detectable in the blood.


Males are more susceptible to this injury. Clamp time would also be 22 minutes rather than 34 minutes for equivalent injury (the paper focuses more on females)
 
 
Cell Types to Focus On:

Injured Proximal Tubule cells - primary injury target

Macrophages - key immune responder at 12 hours

T cells - secondary immune response

Proximal Tubule Segments 1-3 (PTS1, PTS2, PTS3) - to see healthy vs injured tubule changes

Fibroblasts - early fibrotic signals


```{r}
library(STdeconvolve)
library(ggplot2)
library(patchwork)
library(ggrepel)
library(dplyr)
library(scatterbar)
library(pheatmap)
library(clusterProfiler)
library(org.Mm.eg.db)
library(SpatialCellChat)
library(CellChat)
library(presto)
library(readr)
library(ggpubr)
```


```{r}
sham <- read.csv('C:/Users/lilli/Desktop/visium_male_kidney_shamecontrol.csv.gz')
hour12 <- read.csv('C:/Users/lilli/Desktop/visium_male_kidney_AKI_12hrecovery.csv.gz')
hour4 <- read.csv('C:/Users/lilli/Desktop/visium_male_kidney_AKI_4hrecovery.csv.gz')
```

```{r}
pos_sham <- sham[, c('X','Y')]
gexp_sham <- sham[, 4:ncol(sham)]
pos_hour12<- hour12[, c('X','Y')]
gexp_hour12 <- hour12[, 4:ncol(hour12)]
pos_hour4<- hour4[, c('X','Y')]
gexp_hour4 <- hour4[, 4:ncol(hour4)]
```



```{r}
#gexp is in spots x genes format; cleanCounts expects genes x spots, so we transpose.
counts_sham <- cleanCounts(t(gexp_sham), min.lib.size = 100, verbose = T, plot = T)
counts_hour12 <- cleanCounts(t(gexp_hour12), min.lib.size = 100, verbose = T, plot = T)
counts_hour4 <- cleanCounts(t(gexp_hour4), min.lib.size = 100, verbose = T, plot = T)
```


```{r}
#feature selection for genes
# corpus will be a matrix with genes as rows and spots as columns.
corpus_sham <- restrictCorpus(counts_sham, removeAbove = 1.0, removeBelow = 0.05, nTopOD = 1000)
corpus_hour12 <- restrictCorpus(counts_hour12, removeAbove = 1.0, removeBelow = 0.05, nTopOD = 1000)
corpus_hour4 <- restrictCorpus(counts_hour4, removeAbove = 1.0, removeBelow = 0.05, nTopOD = 1000)
```


```{r}
#check to see how many clusters to use
fitLDA(t(as.matrix(corpus_sham)), Ks = seq(2, 9, by = 1),
               perc.rare.thresh = 0.05,
               plot=T,
               verbose=T)
fitLDA(t(as.matrix(corpus_hour4)), Ks = seq(2, 9, by = 1),
               perc.rare.thresh = 0.05,
               plot=T,
               verbose=T)
fitLDA(t(as.matrix(corpus_hour12)), Ks = seq(2, 9, by = 1),
               perc.rare.thresh = 0.05,
               plot=T,
               verbose=T)
```


```{r}
# Rename columns using dplyr
pos_sham <- pos_sham %>% rename(x=X, y=Y)
pos_hour12 <- pos_hour12 %>% rename(x=X, y=Y)
pos_hour4 <- pos_hour4 %>% rename(x=X, y=Y)
```


```{r}
#fit LDA models across a range of K values (number of cell types)
#fitLDA expects spots in rows and genes in columns, so we transpose corpus.
ldas_sham <- fitLDA(t(as.matrix(corpus_sham)), Ks = 7) 

#opting for K = 7 
optLDA_sham <- optimalModel(models = ldas_sham, opt = '7')

#extract deconvolved cell-type proportions (theta) and transcriptional profiles (beta)
results_sham <- getBetaTheta(optLDA_sham, perc.filt = 0.05, betaScale = 1000)
deconProp_sham <- results_sham$theta  # Spots x cell types (clusters)
deconGexp_sham <- results_sham$beta   # Genes x cell types (clusters)

#scatterbar to create a scattered stacked bar chart 
p_deconvolve_sham <- scatterbar(
  data         = as.data.frame(deconProp_sham),
  xy           = pos_sham,
  size_x       = 10,       # adjust to control bar width in x dimension
  size_y       = 10,       # adjust to control bar height in y dimension
  padding_x    = 0.01,     # add some padding to the x axis
  padding_y    = 0.01,     # add some padding to the y axis
  show_legend  = T,
  legend_title = 'Cell Types',
  verbose      = T) +
  labs(x = 'X', y = 'Y') +
  ggtitle('Deconvolved Cell-Type Proportions (k=7)') +
  theme_classic()

p_deconvolve_sham
```

```{r}
#fit LDA models across a range of K values (number of cell types)
#fitLDA expects spots in rows and genes in columns, so we transpose corpus.
ldas_hour12 <- fitLDA(t(as.matrix(corpus_hour12)), Ks = 7) 

#opting for K = 7 
optLDA_hour12 <- optimalModel(models = ldas_hour12, opt = '7')

#extract deconvolved cell-type proportions (theta) and transcriptional profiles (beta)
results_hour12 <- getBetaTheta(optLDA_hour12, perc.filt = 0.05, betaScale = 1000)
deconProp_hour12 <- results_hour12$theta  # Spots x cell types (clusters)
deconGexp_hour12 <- results_hour12$beta   # Genes x cell types (clusters)


#scatterbar to create a scattered stacked bar chart 
p_deconvolve_hour12 <- scatterbar(
  data         = as.data.frame(deconProp_hour12),
  xy           = pos_hour12,
  size_x       = 10,       # adjust to control bar width in x dimension
  size_y       = 10,       # adjust to control bar height in y dimension
  padding_x    = 0.01,     # add some padding to the x axis
  padding_y    = 0.01,     # add some padding to the y axis
  show_legend  = T,
  legend_title = 'Cell Types',
  verbose      = T) +
  labs(x = 'x', y = 'y') +
  ggtitle('Deconvolved Cell-Type Proportions (k=7) 12 Hours') +
  theme_classic()

p_deconvolve_hour12
```

```{r}
#fit LDA models across a range of K values (number of cell types)
#fitLDA expects spots in rows and genes in columns, so we transpose corpus.
ldas_hour4 <- fitLDA(t(as.matrix(corpus_hour4)), Ks = 7) 

#opting for K = 7 
optLDA_hour4 <- optimalModel(models = ldas_hour4, opt = '7')

#extract deconvolved cell-type proportions (theta) and transcriptional profiles (beta)
results_hour4 <- getBetaTheta(optLDA_hour4, perc.filt = 0.05, betaScale = 1000)
deconProp_hour4 <- results_hour4$theta  # Spots x cell types (clusters)
deconGexp_hour4 <- results_hour4$beta   # Genes x cell types (clusters)


#scatterbar to create a scattered stacked bar chart 
p_deconvolve_hour4 <- scatterbar(
  data         = as.data.frame(deconProp_hour4),
  xy           = pos_hour4,
  size_x       = 10,       # adjust to control bar width in x dimension
  size_y       = 10,       # adjust to control bar height in y dimension
  padding_x    = 0.01,     # add some padding to the x axis
  padding_y    = 0.01,     # add some padding to the y axis
  show_legend  = T,
  legend_title = 'Cell Types',
  verbose      = T) +
  labs(x = 'x', y = 'y') +
  ggtitle('Deconvolved Cell-Type Proportions (k=7) 4 Hours') +
  theme_classic()

p_deconvolve_hour4
```

```{r}
#I got this from https://humphreyslab.com/SingleCell/
markers <- list(
  Pod= c('Podxl','Nphs2','Sema3g','Cdkn1c','Rhpn1','Ptpro','Synpo','Sparc','Ccn2','Vim'),
  PTs12= c('Cyp4a14','Slc34a1','Slc13a3','Gpx3','Miox','Aldob','Acaa1b','Slc22a8','Slc22a6','Pck1'),
  PTs3= c('Serpina1f','Cyp2a4','Slc22a19','Akr1c18','Scd1','Slc22a7','Aadat','Kap','Napsa','Lrp2'),
  PTs3_TAL= c('Kap','Serpina1f','Napsa','Scd1','Umod','Aadat','Slc22a7','Havcr1','S100a8','Lgals1'),
  TAL= c('Umod','Slc12a1','Clcnka','Slc14a2','Cd24a','Tspan8','Cldn10','Lcn2','Ckb','Ppp1r1a'),
  DTL= c('Aqp1','Slc14a2','Tinagl1','Akap12','Thbs1','Gprc5a','Nupr1','S100a6','Hspb1','Ifit3'),
  DCT= c('Slc12a3','Pvalb','Calb1','Clcnkb','Klk1','S100g','Slc8a1','Kl','Wnk4','Defb1'),
  PC= c('Aqp2','Muc20','Psca','Wnt7b','Fxyd4','Krt18','Muc1','Gstm2','Akr1b3','Bcam'),
  IC= c('Atp6v1g3','Rhbg','Slc26a4','Rhcg','Hsd11b2','Clu','Car15','Tacstd2','Cldn4','Scnn1b'),
  Uro= c('Upk3a','Upk1a','Upk1b','Upk2','Krt15','Krt5','Krt18','Krt19','Ly6d','Msln'),
  Fib= c('Dcn','Col1a1','Col1a2','Col3a1','Acta2','Myh11','Eln','Bgn','Tagln','Myl9'),
  Immune= c('Lyz2','Ctss','Cd52','Tyrobp','C1qc','C1qa','C1qb','Apoe','Cfh','Cd74', 'Cd68', 'Il6'))
```

```{r}
score_topics <- function(beta_matrix, marker_list) {
  # beta_matrix: topics x genes (rows = topics, cols = genes)
  scores <- sapply(marker_list, function(genes) {
    genes_present <- intersect(genes, colnames(beta_matrix))  # genes are in COLUMNS
    if (length(genes_present) == 0) return(rep(0, nrow(beta_matrix)))
      # average across marker genes per topic
    rowMeans(beta_matrix[, genes_present, drop = F])})
  return(as.data.frame(scores))  # topics x cell_types
}
```


```{r}
annotate_topics <- function(deconGexp, deconProp, condition_name, pos, corpus) {
  
  #align rownames using corpus barcodes
  barcodes <- colnames(corpus)
  pos <- as.data.frame(pos)
  colnames(pos) <- c('x', 'y')
  
  n <- min(nrow(pos), nrow(deconProp))
  
  pos_aligned<- pos[1:n, ]
  deconProp_aligned <- deconProp[1:n, ]
  spot_ids<- barcodes[1:n]
  
  rownames(pos_aligned)<- spot_ids
  rownames(deconProp_aligned)<- spot_ids
  
  
  #score topics against marker lists 
  topic_scores <- score_topics(deconGexp, markers)
  rownames(topic_scores) <- rownames(deconGexp)
  
  cat('\nMax marker score per topic:\n')
  print(round(apply(topic_scores, 1, max), 4))
  
  #assign cell types based on max score
  best_match <- apply(topic_scores, 1, function(x) {
    if (all(x == 0)) return('Unassigned')
    names(which.max(x))})
  #I wanted to make it unique because there might be a reason why STdeconvolve 
  #seperated the clusters
  #they could be more injured cells compared to the other cluster
  best_match_unique <- make.unique(best_match, sep = '_')
  
  cat('\nCell type assignments for', condition_name, '\n')
  print(data.frame(Topic=rownames(topic_scores),
                   CellType=best_match,
                   UniqueLabel = best_match_unique))
  
  #create a heatmap of the topics scores
  topic_scores_clean <- topic_scores[rowSums(topic_scores) > 0, , drop = F]
  
  if (nrow(topic_scores_clean) > 1) {
    pheatmap::pheatmap(topic_scores_clean,
                       scale= 'row',
                       cluster_rows = F,
                       main= paste('Topic–Cell Type Scores:', condition_name),
                       angle_col= 45,)}
  
  #rename deconProp columns 
  deconProp_named <- deconProp_aligned
  colnames(deconProp_named) <- best_match_unique
  
  #remake the scatterbar plot but with the annotation labels
  p <- scatterbar(
    data         = as.data.frame(deconProp_named),
    xy           = pos_aligned,
    size_x       = 10,
    size_y       = 10,
    padding_x    = 0.01,
    padding_y    = 0.01,
    show_legend  = T,
    legend_title = 'Cell Types',
    verbose      = T) +
    labs(x = 'X', y = 'Y') +
    ggtitle(paste('Deconvolved Cell-Type Proportions:', condition_name)) +
    theme_classic()
  
  return(list(scores= topic_scores,
              assignments= best_match_unique,
              deconProp_named = deconProp_named,
              pos_aligned= pos_aligned,
              plot = p))}
```


```{r}
res_sham<-annotate_topics(deconGexp_sham,   deconProp_sham,   'Sham', pos_sham,   corpus_sham)
res_hour4<-annotate_topics(deconGexp_hour4,  deconProp_hour4,  '4h',   pos_hour4,  corpus_hour4)
res_hour12<-annotate_topics(deconGexp_hour12, deconProp_hour12, '12h',  pos_hour12, corpus_hour12)
```


```{r}
#heatmap of average marker gene expression
plot_avg_marker_heatmap <- function(deconGexp, res_obj, condition_name, markers) {
  
  # deconGexp: topics x genes matrix
  # res_obj: result from annotate_topics (contains assignments)
  #get the cell type assignments for each topic
  topic_assignments <- res_obj$assignments
  names(topic_assignments) <- rownames(deconGexp)
  
  
  #get all unique cell types in your markers list
  all_cell_types <- names(markers)
  
  #matrix to store average expression for each topic vs each cell type
  avg_expr_matrix <- matrix(0, nrow=nrow(deconGexp), ncol=length(all_cell_types))
  rownames(avg_expr_matrix) <- rownames(deconGexp)
  colnames(avg_expr_matrix) <- all_cell_types
  
  #for each topic, calculate average expression of markers for each cell type
  for (i in 1:nrow(deconGexp)) {
    topic_name <- rownames(deconGexp)[i]
    
    for (j in 1:length(all_cell_types)) {
      cell_type <- all_cell_types[j]
      marker_genes <- markers[[cell_type]]
      
      #find which marker genes are present in the gene expression matrix
      genes_present <- intersect(marker_genes, colnames(deconGexp))
      
      if (length(genes_present) > 0) {
        #average expression of these marker genes for this topic
        avg_expr_matrix[i, j] <- mean(deconGexp[topic_name, genes_present])}
    }
  }
  
  #scale each row
  avg_expr_scaled <- t(scale(t(avg_expr_matrix)))
  
  #annotation for topics
  topic_annotation <- data.frame(
    Assigned_CellType = factor(topic_assignments[rownames(avg_expr_scaled)]))
  rownames(topic_annotation) <- rownames(avg_expr_scaled)
  
  # Create annotation colors
  #get number of cell types and create colors exactly like scatterbar
  cell_types_in_annotation <- levels(topic_annotation$Assigned_CellType)
  n_celltypes_anno <- length(cell_types_in_annotation)

  #create colors using same method as scatterbar
  celltype_colors <- rainbow(n_celltypes_anno)
  names(celltype_colors) <- cell_types_in_annotation
  
  
  #https://www.rdocumentation.org/packages/pheatmap/versions/1.0.13/topics/pheatmap
  p<- pheatmap::pheatmap(
    avg_expr_scaled,
    cluster_rows = F,
    cluster_cols = F,
    main = paste('Average Marker Expression -', condition_name),
    annotation_row = data.frame(
      CellType = factor(topic_assignments[rownames(avg_expr_scaled)])),
    annotation_colors = list(
      CellType = celltype_colors[unique(topic_assignments)]),
    color = colorRampPalette(c('navy', 'white', 'firebrick3'))(100),
    display_numbers = T,
    width = 10,
    height = 6)
  
  # Also create a simpler version with topic labels including assignments
  rownames_clean <- paste0(rownames(avg_expr_scaled), ' (', topic_assignments, ')')
  avg_expr_named <- avg_expr_scaled
  rownames(avg_expr_named) <- rownames_clean
  
  return(list(avg_matrix = avg_expr_matrix,
              avg_scaled = avg_expr_scaled,
              topic_assignments = topic_assignments,
              plot = p))
}
```

```{r}
avg_sham   <- plot_avg_marker_heatmap(deconGexp_sham, res_sham, 'Sham', markers)
avg_hour4  <- plot_avg_marker_heatmap(deconGexp_hour4, res_hour4, '4h', markers)
avg_hour12 <- plot_avg_marker_heatmap(deconGexp_hour12, res_hour12, '12h', markers)
```

```{r}
#compute mean proportion per cell type for each condition ────────────────
get_mean_props <- function(deconProp_named, condition) {
  as.data.frame(deconProp_named) %>%
    summarise(across(everything(), mean)) %>%
    pivot_longer(everything(), names_to = 'CellType', values_to = 'Proportion') %>%
    mutate(Condition = condition)
}

df_sham<- get_mean_props(res_sham$deconProp_named,   'Sham')
df_hour4<- get_mean_props(res_hour4$deconProp_named,  '4h')
df_hour12<- get_mean_props(res_hour12$deconProp_named, '12h')

all_cell_types <- unique(c(
  unique(df_sham$CellType),
  unique(df_hour4$CellType),
  unique(df_hour12$CellType)
))

#make a complete grid of all conditions and cell types
df_all <- expand.grid(Condition = c('Sham', '4h', '12h'),
                      CellType = all_cell_types,
                      stringsAsFactors = F) %>%
                      left_join(
                        bind_rows(df_sham, df_hour4, df_hour12),
                        by = c('Condition', 'CellType')) %>%
                      #replace NA with 0
                      mutate(
                        Proportion = ifelse(is.na(Proportion), 0, Proportion),
                        Condition = factor(Condition, levels = c('Sham', '4h', '12h')))

ggplot(df_all, aes(x = CellType, y = Proportion, fill = Condition)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c('Sham' = 'green', '4h' = 'orange', '12h' = 'purple')) +
  labs(title = 'Cell type proportions: Sham vs 4h vs 12h',
       x= 'Cell Type',
       y= 'Mean Proportion',
       fill= 'Condition') +
  theme_classic() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        plot.title   = element_text(hjust = 0.5))+
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim=c(0, 0.4))+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))
  

```

```{r}
#compute mean proportion per cell type for each condition ────────────────
get_mean_props <- function(deconProp_named, condition) {
  as.data.frame(deconProp_named) %>%
    summarise(across(everything(), mean)) %>%
    pivot_longer(everything(), names_to = 'CellType', values_to = 'Proportion') %>%
    mutate(Condition = condition)
}

df_sham<- get_mean_props(res_sham$deconProp_named,   'Sham')
df_hour4<- get_mean_props(res_hour4$deconProp_named,  '4h')
df_hour12<- get_mean_props(res_hour12$deconProp_named, '12h')

all_cell_types <- unique(c(
  unique(df_sham$CellType),
  unique(df_hour4$CellType),
  unique(df_hour12$CellType)
))

#make a complete grid of all conditions and cell types
df_all <- expand.grid(Condition = c('Sham', '4h', '12h'),
                      CellType = all_cell_types,
                      stringsAsFactors = F) %>%
                      left_join(
                        bind_rows(df_sham, df_hour4, df_hour12),
                        by = c('Condition', 'CellType')) %>%
                      #replace NA with 0
                      mutate(
                        Proportion = ifelse(is.na(Proportion), 0, Proportion),
                        Condition = factor(Condition, levels = c('Sham', '4h', '12h')))

ggplot(df_all, aes(x = CellType, y = Proportion, fill = Condition)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c('Sham' = 'green', '4h' = 'orange', '12h' = 'purple')) +
  labs(title = 'Cell type proportions: Sham vs 4h vs 12h',
       x= 'Cell Type',
       y= 'Mean Proportion',
       fill= 'Condition') +
  theme_classic() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        plot.title   = element_text(hjust = 0.5))+
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim=c(0, 0.4))+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))

```
```{r}
df_all <- df_all %>%
  group_by(Condition) %>%
  mutate(Tissue_Percent = (Proportion / sum(Proportion)) * 100) %>%
  ungroup()

ggplot(df_all, aes(x = CellType, y = Tissue_Percent, fill = Condition)) +
  geom_bar(stat = 'identity', position = position_dodge(width = 0.8), width = 0.7) +
  scale_fill_manual(values = c('Sham' = 'green', '4h' = 'orange', '12h' = 'purple')) +
  labs(title = 'Cell type proportions: Sham vs 4h vs 12h',
       x = 'Cell Type',
       y = 'Percent of Tissue (%)',
       fill = 'Condition') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        plot.title = element_text(face = "bold", size = 12, hjust = 0.5)) +
  scale_y_continuous(expand = c(0, 0))
```
```{r}
cortex_types  <- c('Pod', 'PTs12', 'PTs3', 'PTs3_TAL', 'DCT')
medulla_types <- c('TAL', 'DTL', 'PC', 'IC')

estimate_cmj <- function(deconProp_named, pos) {
  df <- as.data.frame(deconProp_named)
  
  #find which cortex-specific and medulla-specific cell types exist in our data
  #cortex_types and medulla_types should be defined globally with lists of cell type names
  cortex_cols  <- intersect(cortex_types,  colnames(df))
  medulla_cols <- intersect(medulla_types, colnames(df))
  
  cat('  Cortex cell types found: ', paste(cortex_cols,  collapse=', '), '\n')
  cat('  Medulla cell types found:', paste(medulla_cols, collapse=', '), '\n')
  

  #for each spot, calculate the average proportion of cortex-specific cell types
  #If no cortex types exist, set score to 0
  df$cortex_score  <- if (length(cortex_cols)  > 0) {
    rowMeans(df[, cortex_cols, drop = F]) } else {0}
  
  #for each spot, calculate the average proportion of medulla-specific cell types
  # If no medulla types exist, set score to 0
  df$medulla_score <- if (length(medulla_cols) > 0) {
    rowMeans(df[, medulla_cols, drop = F])} else {0}
  
  #a spot belongs to cortex if its cortex_score is higher than medulla_score
  df$region <- ifelse(df$cortex_score > df$medulla_score, 'cortex', 'medulla')
  
  #Y-coordinate from position data (assuming Y-axis runs from cortex to medulla)
  df$y <- as.data.frame(pos)[, 2]
  
  #num spots were classified into each region
  n_cortex<- sum(df$region == 'cortex')
  n_medulla<- sum(df$region == 'medulla')
  cat('  Spots classified as cortex: ', n_cortex, '\n')
  cat('  Spots classified as medulla:', n_medulla, '\n')
  
  # The CMJ is the boundary where cortex ends and medulla begins
  #we find the lowest cortex spot (maximum Y) and highest medulla spot (minimum Y)
  #I assume that the CMJ is the midpoint between these two
  
  if (n_medulla == 0) {
    #No medulla spots detected (possible if section is all cortex)
    cat('No medulla spots found; using median Y as CMJ estimate\n')
    #use median Y coordinate
    cmj_y <- median(df$y)} 
  else if (n_cortex == 0) {
    #case 2: No cortex spots detected (possible if section is all medulla)
    cat('No cortex spots found; using median Y as CMJ estimate\n')
    cmj_y <- median(df$y)}
  else {
    #Case 3: Both regions present, calculate T boundary
    #we find the lowest cortex spot. This is the last cortex before medulla
    max_cortex_y  <- max(df$y[df$region == 'cortex'])
    #highest medulla spot is the first medulla after cortex
    min_medulla_y <- min(df$y[df$region == 'medulla'])
    #CMJ is the average of these two boundary points
    cmj_y <- mean(c(max_cortex_y, min_medulla_y))}
  
  cat('  Estimated CMJ Y coordinate:', cmj_y, '\n')
  
  
  return(list(cmj_y = cmj_y, df = df))}

prep_cmj_df <- function(deconProp_named, pos, cmj_y, condition, counts, macro_markers) {
  
  df <- as.data.frame(deconProp_named)
  
  #macrophage markers present in this condition's counts
  markers_present <- intersect(macro_markers, rownames(counts))
  
  #mean expression of macrophage markers per spot
  macro_expr <- counts[markers_present, 1:nrow(df), drop = FALSE]
  df$macro_loading <- colMeans(macro_expr)
  
  df %>%
    mutate(dist_CMJ  = pos[, 2] - cmj_y, Condition = condition) %>%
    dplyr::select(dist_CMJ, macro_loading, Condition)
}
```


```{r}
cmj_sham<- estimate_cmj(res_sham$deconProp_named,   res_sham$pos_aligned)
cmj_hour4<- estimate_cmj(res_hour4$deconProp_named,  res_hour4$pos_aligned)
cmj_hour12<- estimate_cmj(res_hour12$deconProp_named, res_hour12$pos_aligned)

#https://www.rndsystems.com/resources/cell-markers/immune-cells/macrophages/macrophage-common-markers
#https://www.antibodies.com/primary-antibodies/cell-markers/immune-cell-markers/macrophage-markers
macro_markers <- c('Lyz2','Ctss','Cd52','Tyrobp','C1qc','C1qa','C1qb',
                   'Apoe','Cfh','Cd74','Csf1r','Adgre1','Cx3cr1','Spp1', 'Cd68')

df_sham<- prep_cmj_df(res_sham$deconProp_named, 
                      res_sham$pos_aligned,
                      cmj_sham$cmj_y, 
                      'Sham',
                      counts_sham,
                      macro_markers)

df_hour4<- prep_cmj_df(res_hour4$deconProp_named,
                       res_hour4$pos_aligned,
                       cmj_hour4$cmj_y,
                       'Recovery 4h',
                       counts_hour4,
                       macro_markers)

df_hour12<- prep_cmj_df(res_hour12$deconProp_named,
                        res_hour12$pos_aligned,
                        cmj_hour12$cmj_y,
                        'Recovery 12h',
                        counts_hour12,
                        macro_markers)

df_all <- bind_rows(df_sham, df_hour4, df_hour12) %>%
  mutate(Condition = factor(Condition, levels = c('Sham', 'Recovery 4h', 'Recovery 12h')))

condition_colors <- c('Sham' = 'green', 'Recovery 4h'= 'orange', 'Recovery 12h' = 'purple')

ggplot(df_all, aes(x = dist_CMJ, y = macro_loading, color = Condition, fill = Condition)) +
  geom_smooth(method = 'loess', span = 0.4, se = T, alpha = 0.2, linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = 'dashed', color = 'red', linewidth = 0.6) +
  annotate('text', x = 0, y = Inf, label = 'CMJ', color = 'red', vjust = 1.5, hjust = -0.2, size = 3.5) +
  annotate('text', x = -1500, y = -Inf, label = 'Cortex', color = 'black', vjust = 2.5, size = 3.5) +
  annotate('text', x = 1500, y = -Inf, label = 'Medulla', color = 'black', vjust = 2.5, size = 3.5) +
  scale_color_manual(values = condition_colors) +
  scale_fill_manual(values  = condition_colors) +
  scale_x_continuous(breaks = seq(-2000, 3000, by = 1000), 
                     labels = paste0(seq(-2000, 3000, by = 1000), 'µm')) +
  coord_cartesian(clip = 'off') +
  labs(title = 'Macrophage Expression vs Distance from CMJ',
       subtitle= 'Macrophages Travel From the Cortex to the Medulla Over Time',
       x = 'Distance from Corticomedullary Junction (µm)',
       y = 'Mean Macrophage Marker Expression',
       color = NULL, fill = NULL) +
  scale_y_continuous(expand = c(0, 0)) +
  coord_cartesian(ylim=c(0, 40), clip = 'off')+
  theme_classic()+
  theme(plot.margin= margin(10, 10, 25, 10), axis.title.x= element_text(margin = margin(t = 20)),)+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5), 
        plot.subtitle = element_text(size=10, face='italic', color='grey30', hjust = 0.5))

```


```{r}
df_cmj <- bind_rows(
  df_sham %>% mutate(Condition = 'Sham'),
  df_hour4 %>% mutate(Condition = 'Recovery 4h'), 
  df_hour12 %>% mutate(Condition = 'Recovery 12h')) %>%
  mutate(Condition = factor(Condition, levels = c('Sham', 'Recovery 4h', 'Recovery 12h')))

#pairwise wilcox test
cat('\nPairwise Wilcoxon Tests (Bonferroni corrected)\n')
wilcox_result <- pairwise.wilcox.test(df_cmj$macro_loading, df_cmj$Condition, p.adjust.method = 'bonferroni')
print(wilcox_result)

ggplot(df_cmj, aes(x = Condition, y = macro_loading, fill = Condition)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.3, size = 0.5) +
  scale_fill_manual(values = c('Sham' = 'green', 
                                'Recovery 4h' = 'orange', 
                                'Recovery 12h' = 'purple')) +
  theme_classic() +
  labs(title = 'Macrophage Score by Condition', y = 'Mean Macrophage Marker Expression') +
  stat_compare_means(comparisons = list(c('Sham', 'Recovery 4h'),
                                        c('Sham', 'Recovery 12h'),
                                        c('Recovery 4h', 'Recovery 12h')),
                     method = 'wilcox.test', 
                     label = 'p.signif')+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))
```


```{r}
get_macro_score <- function(counts, deconProp_named) {
  n <- nrow(deconProp_named)
  #macrophage markers actually present in our count data
  markers_present <- intersect(macro_markers, rownames(counts))
  cat('Markers found:', paste(markers_present, collapse = ', '), '\n')
  #mean expression across all macrophage markers for each spot
  #if only 1 marker found, just return that one marker's expression
  macro_expr<- as.matrix(counts[markers_present, 1:n, drop = F])
  if (nrow(macro_expr) == 1) as.numeric(macro_expr[1, ]) else colMeans(macro_expr)
}

macro_sham   <- get_macro_score(counts_sham,   res_sham$deconProp_named)
macro_hour4  <- get_macro_score(counts_hour4,  res_hour4$deconProp_named)
macro_hour12 <- get_macro_score(counts_hour12, res_hour12$deconProp_named)

#correlate macrophage score with each cell type proportion
correlate_with_macro <- function(deconProp_named, macro_score, condition) {
  
  df <- as.data.frame(deconProp_named)
  
  cor_results <- sapply(colnames(df), function(ct) {cor(df[[ct]], macro_score, method = 'spearman')})
  
  data.frame(CellType = names(cor_results),
            Spearman_r = cor_results,
            Condition = condition) %>% 
            arrange(desc(Spearman_r))}

cor_sham <- correlate_with_macro(res_sham$deconProp_named,   macro_sham,   'Sham')
cor_hour4 <- correlate_with_macro(res_hour4$deconProp_named,  macro_hour4,  '4h')
cor_hour12 <- correlate_with_macro(res_hour12$deconProp_named, macro_hour12, '12h')


df_all <- bind_rows(cor_sham, cor_hour4, cor_hour12) %>%
  mutate(Condition = factor(Condition, levels = c('Sham', '4h', '12h')))

ggplot(df_all, aes(x = reorder(CellType, Spearman_r), y = Spearman_r, fill = Spearman_r > 0)) +
  geom_bar(stat = 'identity', width = 0.7) +
  geom_hline(yintercept = 0, linetype = 'dashed', color = 'black') +
  scale_fill_manual(values = c('TRUE' = '#E74C3C', 'FALSE' = '#3498DB'), guide = 'none') +
  facet_wrap(~ Condition, ncol = 3) +
  coord_flip() +
  labs(title = 'Correlation of cell type proportion with macrophage marker expression',
       x = NULL,
       y = 'Spearman correlation') +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5),
        strip.text  = element_text(size = 11, face = 'bold'),
        axis.text.y = element_text(size = 10))

#scatter plot for top correlated cell type in 12h 
top_ct_12h <- cor_hour12$CellType[1]
cat('\nTop correlated cell type in 12h:', top_ct_12h, '\n')

df_scatter <- data.frame(macro_score = macro_hour12, ct_prop = res_hour12$deconProp_named[, top_ct_12h])

ggplot(df_scatter, aes(x = ct_prop, y = macro_score)) +
  geom_point(alpha = 0.3, size = 0.8, color = '#E74C3C') +
  geom_smooth(method = 'lm', color = 'black', linewidth = 0.8) +
  labs(title = paste('Macrophage score vs', top_ct_12h, 'proportion (12h)'),
       x = paste(top_ct_12h, 'topic proportion'),
       y = 'Mean macrophage marker expression') +
  theme_classic() +
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))
```


```{r}
#get_top_genes: rows are topics, cols are genes 
get_top_genes <- function(deconGexp, topic_row, n_genes = 100) {
  #topic_row is the numeric rowname
  topic_vec <- deconGexp[topic_row, ]
  top_genes <- names(sort(topic_vec, decreasing = T))[1:n_genes]
  return(top_genes)
}

#run_enrichment using topic number + cell type label
run_enrichment <- function(deconGexp, topic_row, cell_type_label, 
                           n_genes = 100, condition_name = '') {
  
  top_genes <- get_top_genes(deconGexp, topic_row, n_genes)
  #Following the turtorial below
  #https://bioinformatics.ccr.cancer.gov/docs/btep-coding-club/CC2023/FunctionalEnrich_clusterProfiler/
  #SYMBOL: Human-readable gene names 
  #ENTREZID: NCBI's database IDs (required for KEGG and GO)
  #org.Mm.eg.db is the mouse annotation database
  entrez <- bitr(top_genes,
                 fromType= 'SYMBOL',
                 toType= 'ENTREZID',
                 OrgDb= org.Mm.eg.db)
  
  if (nrow(entrez) == 0) stop('No Entrez IDs found')
  
  #GO: what biological processes are over-represented
  #tests if gene list has more genes from certain GO terms than expected by chance
  ego <- enrichGO(gene= entrez$ENTREZID,
                  OrgDb = org.Mm.eg.db,
                  ont = 'BP',
                  pAdjustMethod = 'BH',
                  pvalueCutoff  = 0.05,
                  qvalueCutoff  = 0.05,
                  readable = T)
  
  #KEGG Pathway Enrichment Analysis
  #tests if genes are enriched in specific signaling/metabolic pathways
  ekegg <- enrichKEGG(gene = entrez$ENTREZID,
                      organism = 'mmu',
                      pAdjustMethod = 'BH',
                      pvalueCutoff  = 0.05)
  
  return(list(GO= ego,KEGG = ekegg, genes = top_genes, label = paste(condition_name, cell_type_label)))
}

run_all_enrichments <- function(deconGexp, deconProp_named, condition_name, n_genes = 100) {
  
  cell_type_labels <- colnames(deconProp_named) 
  topic_rows <- rownames(deconGexp)     
  
  #we assume that in the same order — topic 1 = label 1 etc.
  #lapply iterates over each index i from 1 to the number of cell types
  results <- lapply(seq_along(cell_type_labels), function(i) {
    topic_row <- topic_rows[i]
    cell_type_label <- cell_type_labels[i]
    cat('Running enrichment for:', condition_name, '-', cell_type_label, 
        '(Topic', topic_row, ')\n')
    run_enrichment(deconGexp, topic_row, cell_type_label, n_genes, condition_name)})
  
  names(results) <- cell_type_labels
  return(results)
}

enrich_sham   <- run_all_enrichments(deconGexp_sham,   res_sham$deconProp_named,   'Sham')
enrich_hour4  <- run_all_enrichments(deconGexp_hour4,  res_hour4$deconProp_named,  '4h')
enrich_hour12 <- run_all_enrichments(deconGexp_hour12, res_hour12$deconProp_named, '12h')

```

```{r}
plot_enrichment <- function(enrich_df, top_n = 10, title = NULL) {
  df <- enrich_df %>%
    arrange(p.adjust) %>%
    head(top_n) %>%
    mutate(log10p= -log10(p.adjust), 
           Description = factor(Description, levels = rev(Description)))
  
  ggplot(df, aes(x = log10p, y = Description)) +
    geom_bar(stat = 'identity', fill = '#1f77b4', width = 0.7) +
    geom_vline(xintercept = -log10(0.05), linetype = 'dashed', color = 'red', linewidth = 0.5) +
    labs(title = title, x = '-log10(adjusted p-value)', y = NULL) +
    theme_minimal() +
    theme(plot.title = element_text(hjust = 0.5, size = 11),
          axis.text.y= element_text(size = 9),
          panel.grid.major.y = element_blank())
}

plot_all_topics <- function(enrich_results, result_type = 'GO') {
  for (topic in names(enrich_results)) {
    res <- enrich_results[[topic]]
    if (is.null(res)) {
      cat('Skipped (NULL):', topic, '\n')
      next}
    
    df <- as.data.frame(res[[result_type]])
    if (nrow(df) == 0) {
      cat('No significant terms for:', topic, '\n')
      next}
    
    p <- plot_enrichment(df, title = paste(result_type, ':', res$label))
    print(p)
    cat('Plotted:', topic, '\n')}
}

```

```{r}
plot_all_topics(enrich_sham,   result_type = 'GO')
plot_all_topics(enrich_hour4,  result_type = 'GO')
plot_all_topics(enrich_hour12, result_type = 'GO')
```

```{r}
plot_all_topics(enrich_sham,   result_type = 'KEGG')
plot_all_topics(enrich_hour4,  result_type = 'KEGG')
plot_all_topics(enrich_hour12, result_type = 'KEGG')
```


```{r}
normalize_counts <- function(counts) {
  counts_norm <- sweep(counts, 2, colSums(counts), '/') * 10000
  log1p(counts_norm)
}

#STdeconvolve does not use normalized counts
counts_sham_norm   <- normalize_counts(counts_sham)
counts_hour4_norm  <- normalize_counts(counts_hour4)
counts_hour12_norm <- normalize_counts(counts_hour12)

prep_meta <- function(deconProp_named) {
  labels <- colnames(deconProp_named)[apply(deconProp_named, 1, which.max)]
  data.frame(cell_type = labels, row.names = rownames(deconProp_named))}

meta_sham<- prep_meta(res_sham$deconProp_named)
meta_hour4 <- prep_meta(res_hour4$deconProp_named)
meta_hour12 <- prep_meta(res_hour12$deconProp_named)

#create CellChat objects
create_chat <- function(counts_norm, meta, n, condition) {
  chat <- createCellChat(object = counts_norm[, 1:n],
                         meta = meta,
                         group.by = 'cell_type')
  cat('Created CellChat for:', condition, '| Groups:', levels(chat@idents), '\n')
  return(chat)
}

chat_sham <- create_chat(counts_sham_norm, meta_sham, nrow(meta_sham), 'Sham')
chat_hour4 <- create_chat(counts_hour4_norm, meta_hour4, nrow(meta_hour4), '4h')
chat_hour12 <- create_chat(counts_hour12_norm, meta_hour12, nrow(meta_hour12), '12h')

CellChatDB.use <- subsetDB(CellChatDB.mouse,
                            search = c('Secreted Signaling', 'ECM-Receptor', 'Cell-Cell Contact'),
                            non_protein = F)

#I followed this tutorial: https://deepwiki.com/sqjin/CellChat/6.1-basic-usage-tutorial
#https://deepwiki.com/sqjin/CellChat/6.3-spatial-data-analysis-tutorial
run_chat <- function(chat, condition) {
  cat('\nRunning CellChat for:', condition, '\n')
  
  chat@DB <- CellChatDB.use
  chat<- subsetData(chat)
  chat <- identifyOverExpressedGenes(chat)
  chat <- identifyOverExpressedInteractions(chat)
  chat <- computeCommunProb(chat, type = 'triMean')
  chat <- filterCommunication(chat, min.cells = 10)
  chat <- computeCommunProbPathway(chat)
  chat <- aggregateNet(chat)
  chat <- netAnalysis_computeCentrality(chat, slot.name = 'netP')
  return(chat)
}

chat_sham <- run_chat(chat_sham,   'Sham')
chat_hour4 <- run_chat(chat_hour4,  '4h')
chat_hour12 <- run_chat(chat_hour12, '12h')

#save because it takes a long time to run
readr::write_rds(chat_sham,   'chat_sham.rds')
readr::write_rds(chat_hour4,  'chat_hour4.rds')
readr::write_rds(chat_hour12, 'chat_hour12.rds')
```

```{r}
#rename the TAL_1 to injured 
colnames(res_hour12$deconProp_named)[colnames(res_hour12$deconProp_named) == 'TAL_1'] <- 'Injured_TAL'

#rebuild chat_hour12 with updated labels 
#use the renamed deconProp (with Injured_TAL) to rebuild meta
meta_hour12_updated <- prep_meta(res_hour12$deconProp_named)
cat('Updated groups:', unique(meta_hour12_updated$cell_type), '\n')

#rebuild and rerun CellChat
chat_hour12 <- create_chat(counts_hour12_norm, meta_hour12_updated, nrow(meta_hour12_updated), '12h')
chat_hour12 <- run_chat(chat_hour12, '12h')

df_net <- subsetCommunication(chat_hour12)
cat('Number of significant interactions:', nrow(df_net), '\n')
cat('Signaling sources:', unique(df_net$source), '\n')
cat('Signaling targets:', unique(df_net$target), '\n')

#since there is no Immune in 12h, look at what Injured_TAL sends
#Find which cell types are receiving signals from Injured_TAL
df_from_injured_tal <- df_net %>%
  filter(source == 'Injured_TAL') %>%
  arrange(desc(prob))

cat('\nSignals from Injured_TAL:\n')
print(head(df_from_injured_tal, 20))
```



```{r}
#Immune recruitment L-R pairs
#Gray, A. L., Karlsson, et. al. (2023).  Chemokine CXCL4 interactions with extracellular matrix proteoglycans mediate widespread immune cell recruitment independent of chemokine receptors. Cell Reports, 42(1), 111930. https://doi.org/10.1016/j.celrep.2022.111930
#Zhang, J., Peng, Q., Fan, J. et al. Single-cell and spatial transcriptomics reveal SPP1-CD44 signaling drives primary resistance to immune checkpoint inhibitors in RCC. J Transl Med 22, 1157 (2024). https://doi.org/10.1186/s12967-024-06018-5
#Yogesh Saini, Owais Mohammad, Brandon W Lewis, Razia Sultana, Understanding macrophage recruitment to the stressed airspaces, The Journal of Immunology, Volume 196, Issue 1_Supplement, May 2016, Page 60.10, https://doi.org/10.4049/jimmunol.196.Supp.60.10
#Li, M., Wang, M., Wen, Y., Zhang, H., Zhao, G. N., & Gao, Q. (2023). Signaling pathways in macrophages: molecular mechanisms and therapeutic targets. MedComm, 4(5), e349. https://doi.org/10.1002/mco2.349

#macrophage recruitment pathways (from literature)
recruitment_pathways <- c('CCL', 'CXCL', 'CX3C', 'CCL', 'CCR', 'CSF', 'CSF1R', 'MIF', 'SPP1',                                         'COMPLEMENT', 'TGFb', 'IL1', 'IL6', 'IL10', 'IL34', 'TNF', 'ANGPTL', 
                          'ANGPT',  'VEGF', 'TREM1', 'TREM2', 'SOX9')


#find which of these are present in data
present_recruitment_sham <- intersect(recruitment_pathways, chat_sham@netP$pathways)
present_recruitment_hour4 <- intersect(recruitment_pathways, chat_hour4@netP$pathways)
present_recruitment_hour12 <- intersect(recruitment_pathways, chat_hour12@netP$pathways)
cat('Recruitment pathways present:', paste(present_recruitment_sham, collapse=', '), '\n')
cat('Recruitment pathways present:', paste(present_recruitment_hour4, collapse=', '), '\n')
cat('Recruitment pathways present:', paste(present_recruitment_hour12, collapse=', '), '\n')

#What is signalling to Immune cells?
netVisual_bubble(chat_hour4,
                 sources.use = setdiff(levels(chat_hour4@idents), 'Immune'),
                 targets.use= 'Immune',
                 signaling = present_recruitment_hour4,
                 remove.isolate = T,
                 title.name = 'Recruitment signals Immune (4h)')

netVisual_bubble(chat_sham,
                 sources.use = setdiff(levels(chat_sham@idents), 'Immune'),
                 targets.use = 'Immune',
                 signaling = present_recruitment_sham,
                 remove.isolate = T,
                 title.name = 'Recruitment signals  Immune (Sham)')

netVisual_bubble(chat_hour12,
                sources.use = 'Injured_TAL',
                targets.use = levels(chat_hour12@idents),
                signaling = present_recruitment_hour12,  
                remove.isolate = T,
                title.name     = 'Macrophage Recruitment Signals from Injured_TAL (12h)',
                font.size = 12
)
```


```{r}
plot_spatial_gene <- function(gene, counts_norm, pos_aligned, condition) {
  n <- nrow(pos_aligned)
  pos_df <- as.data.frame(pos_aligned)
  colnames(pos_df) <- c('x', 'y')
  
  pos_df$expr <- as.numeric(counts_norm[gene, 1:n])
  
  max_val <- round(max(pos_df$expr,  na.rm=T), 2)
  mean_val <- round(mean(pos_df$expr, na.rm=T), 3)
  
  ggplot(pos_df, aes(x = x, y = y, color = expr)) +
    geom_point(size = 1.5) +
    scale_color_gradientn(colors = c('blue', 'white',  'red'),
                          values  = c(0, 0.5, 1),
                          name   = 'log\n expr') +
    labs(title = paste0(gene, ' - ', condition,
                        '\nmax=', max_val, ' | mean=', mean_val)) +
    theme_classic()+
    theme(plot.title = element_text(face = "bold", size =30, hjust = 0.5))
    
}


p1 <- plot_spatial_gene('Spp1', counts_sham_norm, res_sham$pos_aligned, 'Sham')
p2 <- plot_spatial_gene('Spp1', counts_hour4_norm,  res_hour4$pos_aligned,  '4h')
p3 <- plot_spatial_gene('Spp1', counts_hour12_norm, res_hour12$pos_aligned, '12h')
```

```{r}
com<- p1 + p2 +p3
```

```{r}
ggsave('spp1.png', com,
       width = 30, height = 10, dpi = 100)
```

```{r}
#SPP1 is the dominant signal, plot across all conditions
#network graph
par(mfrow = c(1, 3))
netVisual_aggregate(chat_sham,   signaling = 'SPP1', layout = 'circle'); title('SPP1 - Sham')
netVisual_aggregate(chat_hour4,  signaling = 'SPP1', layout = 'circle'); title('SPP1 - 4h')
netVisual_aggregate(chat_hour12, signaling = 'SPP1', layout = 'circle'); title('SPP1 - 12h')
par(mfrow = c(1, 1))
```


```{r}
#network graph for repair
par(mfrow = c(1, 3))
netVisual_aggregate(chat_sham,   signaling = 'TGFb', layout = 'circle'); title('SPP1 - Sham')
netVisual_aggregate(chat_hour4,  signaling = 'TGFb', layout = 'circle'); title('SPP1 - 4h')
netVisual_aggregate(chat_hour12, signaling = 'TGFb', layout = 'circle'); title('SPP1 - 12h')
par(mfrow = c(1, 1))
```

```{r}
p4 <- plot_spatial_gene('Tgfb1', counts_sham_norm, res_sham$pos_aligned, 'Sham')
p5 <- plot_spatial_gene('Tgfb1', counts_hour4_norm,  res_hour4$pos_aligned,  '4h')
p6 <- plot_spatial_gene('Tgfb1', counts_hour12_norm, res_hour12$pos_aligned, '12h')
```


```{r}
comp<- p4 + p5 +p6
```

```{r}
ggsave('tgfb.png', comp,
       width = 30, height = 10, dpi = 100)
```


```{r}
p4 <- plot_spatial_gene('Sox9', counts_sham_norm, res_sham$pos_aligned, 'Sham')
p5 <- plot_spatial_gene('Sox9', counts_hour4_norm,  res_hour4$pos_aligned,  '4h')
p6 <- plot_spatial_gene('Sox9', counts_hour12_norm, res_hour12$pos_aligned, '12h')
```



```{r}
#pathway strengths per condition
get_pathway_strength <- function(chat, condition) {
  
  df <- subsetCommunication(chat, slot.name = 'netP') 
  
  df %>%
    group_by(pathway_name) %>%
    summarise(total_prob = sum(prob), .groups = 'drop') %>%
    arrange(desc(total_prob)) %>%
    mutate(Condition = condition)
}

strength_sham   <- get_pathway_strength(chat_sham, 'Sham')
strength_hour4  <- get_pathway_strength(chat_hour4, '4h')
strength_hour12 <- get_pathway_strength(chat_hour12, '12h')

df_strength <- rbind(strength_sham, strength_hour4, strength_hour12) %>%
  mutate(Condition = factor(Condition, levels = c('Sham','4h','12h')))


top_pathways <- df_strength %>%
  group_by(pathway_name) %>%
  summarise(max_prob = max(total_prob)) %>%
  arrange(desc(max_prob)) %>%
  slice_head(n = 20) %>%
  pull(pathway_name)

df_strength %>%
  filter(pathway_name %in% top_pathways) %>%
  ggplot(aes(x = reorder(pathway_name, total_prob), 
             y = total_prob, fill = Condition)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = c('Sham'='green',
                                '4h'='orange',
                                '12h'='purple')) +
  coord_flip(ylim = c(0, 80)) +
  scale_y_continuous(expand = c(0, 0)) + 
  labs(title = 'Pathway information flow across IRI timepoints',
       subtitle= expression(italic('ECM repair and Immmune Cell Recuitment Over Time')),
       x = NULL, y = 'Total communication probability') +
  theme_classic()+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5), 
        plot.subtitle = element_text(size=10, face='italic', color='grey30', hjust = 0.5))
```


```{r}
ggplot(sham, aes(x=X, y=Y, color=Cd8a))+geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour12, aes(x=X, y=Y, color=Cd8a)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour4, aes(x=X, y=Y, color=Cd8a)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
```


```{r}
p1 <- ggplot(sham, aes(x = X, y = Y, color = Cd68)) +
  geom_point() +
  theme_classic() +
  scale_color_gradient(limits = c(0, 10), low = 'blue', high = 'red') +
  labs(title = 'Sham: CD68 Expression')+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))
p2 <- ggplot(hour12, aes(x = X, y = Y, color = Cd68)) +
  geom_point() +
  theme_classic() +
  scale_color_gradient(limits = c(0, 10), low = 'blue', high = 'red') +
  labs(title = '12h: CD68 Expression')+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))

p3 <- ggplot(hour4, aes(x = X, y = Y, color = Cd68)) +
  geom_point() +
  theme_classic() +
  scale_color_gradient(limits = c(0, 10), low = 'blue', high = 'red') +
  labs(title = '4h: CD68 Expression')+
  theme(plot.title = element_text(face = "bold", size = 12, hjust = 0.5))


combined<-p1 +p3 +  p2
```

```{r}
ggsave('macro.png', combined,
       width = 15, height = 5, dpi = 100)
```

```{r}
ggplot(sham, aes(x=X, y=Y, color=Spp1))+geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour12, aes(x=X, y=Y, color=Spp1)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour4, aes(x=X, y=Y, color=Spp1)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
```

```{r}
ggplot(sham, aes(x=X, y=Y, color=Cd3e))+geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour12, aes(x=X, y=Y, color=Cd3e)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
```

```{r}
ggplot(sham, aes(x=X, y=Y, color=Il6))+geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour12, aes(x=X, y=Y, color=Il6)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour4, aes(x=X, y=Y, color=Il6)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
```

```{r}
ggplot(sham, aes(x=X, y=Y, color=Mrc1))+geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour12, aes(x=X, y=Y, color=Mrc1)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
ggplot(hour4, aes(x=X, y=Y, color=Mrc1)) +geom_point()+theme_classic() + scale_color_gradient(low = 'blue', high = 'red',)
```



